---
title: "Exploring the BRFSS data"
output: 
  html_document: 
    fig_height: 4
    highlight: pygments
    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
```

### Load data

```{r load-data}
load("brfss2013.Rdata")
```

* * *

## Part 1: Data
Data is collected by phone, uniformly at random among landlord adults (>= 18 years old) in the US. Thus the study excludes homeless people, who may be especially concerned with risk behaviors. Additional data was added in 2013 for young people. Moreover, I guess that only people who responded to phone are taken into account. Thus the study is not generalizable to the whole population, but only to a subset of it.

Since there is no random assignment, this is an observational study and no causality can be proven, only correlation.

* * *

## Part 2: Research questions

**Research quesion 1:**
Is there any correlation between chccopd1 (Chronic Obstructive Pulmonary Disease) and smokday2 (Frequency Of Days Now Smoking)?

**Research quesion 2:**
Is there any correlation between sleep time (sleptim1) of men and women (based on the sex variable) and the number of children (children)?

**Research quesion 3:**
How are physical exercise (exerany2) and mental/physical health related?

* * *

## Part 3: Exploratory data analysis

**Research quesion 1:**

Let's see the number and percentage of Chronic Obstructive Pulmonary Disease among smokers and non-smokers:
```{r}
table(brfss2013$smokday2, brfss2013$chccopd1)
```
The amount of data collected seems enough to be significant.

```{r}
prop.table(table(brfss2013$smokday2, brfss2013$chccopd1), margin = 1)
```

Among the data collected, occasional smokers have ≈14.6% chance to have Chronic Obstructive Pulmonary Disease whereas non-smokers only have ≈12%, resulting in an increase of 100*(14.6-12)/12 ≈ 21.6%, which is significant. Comparing regular smokers to non-smokers is even more indicative: ≈18% against ≈12, i.e. an increase of 50% in risks.
Therefore, the data seem to indicate that there is indeed a correlation between smoking and higher risk of Chronic Obstructive Pulmonary Disease.

This is also confirmed visually by the following percentage bar plot:
```{r}
ggplot(brfss2013, aes(x=smokday2, fill = chccopd1)) +
  geom_bar(position="fill")
```

**Research question 2:**
Let's see some statistics about sleep time:
```{r}
summary(brfss2013$sleptim1)
```

We can see that some people reported very unusual sleep time (450 hours per day...), we remove those anomalies:
```{r}
brfss2013 <- subset(brfss2013, sleptim1 <= 24)
```

```{r}
t <- table(brfss2013$sleptim1, brfss2013$children)
d <- data.frame("Sleep" = brfss2013$sleptim1, "Children" = brfss2013$children, "Sex" = brfss2013$sex)
d <- subset(d, Children < 10) # we remove the very high number of children since there are uninformative (too few of data)
d <- (d %>% group_by(Children, Sex) %>% summarise(MeanSleep = mean(Sleep), n = n(), .groups = 'drop'))
summary(d)
ggplot(d, aes(x=Children, y=MeanSleep, color = Sex)) +
  geom_line()

```
This curve tends to indicate that having children reduces someone's sleep time, especially with the first child. For a high number of childs, there is not enough data. Men and women have similar sleep duration, but women seems to sleep slighty more.


**Research quesion 3:**
Let's compare the average number of days in bad physical/mental health conditions: 
```{r}
df <- brfss2013 %>% group_by(exerany2) %>% summarise(PhysHealth = mean(physhlth, na.rm=TRUE), MentalHealth = mean(menthlth, na.rm=TRUE), n = n(), .groups = 'drop')
df
```
```{r}
ggplot(df, aes(x=exerany2, y = PhysHealth)) + geom_bar(stat="identity")
```
```{r}
ggplot(df, aes(x=exerany2, y = MentalHealth)) + geom_bar(stat="identity")
```

The absence of physical exercise is highly correlated with bad physical health and mental health. However it is difficult to conclude about any causal possibility: maybe bad physical health cause people to do less exercise, maybe less exercice means worse health, or maybe there is correlation without causation.
